Human Factors Influencing Visual Statistical Inference

### 2014-10-20 sociology_paper.tex
small edits throught the first section

The lineup protocol introduced in \citet{buja:2009} can be used to test the significance of findings during the exploratory data analysis. The methodology is a part of what is called visual statistical inference.  These concepts have been developed further by \citet{majumder:2013} who refined the terminology and validated the lineup protocol with a head to head comparison with conventional inference. One of the major contributions of  \citet{majumder:2013}  is to define the power of the visual test and provide methods to obtain the power for a particular lineup. It was observed that the power can be as good or better than that of a conventional test in some scenarios.

@@

Figure~\ref{fig:tower} displays a lineup of 20 plots where one of the plots is observed data, while the remaining 19 plots are showing from data generated under a null model. Which one of the 20 plots is the most different from the others? 
When asked this question, 12 \hh{out} of 72 observers picked the  plot \hh{of the observed data} (located in spot number $3^2+4$ in the lineup).  The corresponding $p$-value is \hh{0.0077}, indicating sufficient evidence to reject the null hypothesis. 
% , with reasons given being 'asymmetry' (36\%), 'trend' (26\%) or `outliers' (16\%).

\hh{How do we interpret this finding, though?} For that, we need to know the context of the data and we need to have more information about the \hh{how null plots were generated}. 
This particular example investigates the results from the 2012 US presidential election in comparison to the poll results just prior to the election. 
(Although this example is more simplistic than most of the tests conducted to date, it will serve the purpose of illustrating the lineup protocol.) The data is looking at the difference in poll results between the two (major) presidential candidates, Obama and Romney, for all states. Each panel in Figure~\ref{fig:elect-1} shows an `electoral building' where each state in the union is represented by a rectangle. This difference is plotted horizontally, and the height of each box corresponds to the state's electoral votes. Color indicates party affiliation.  The null hypothesis is ``that the election results were consistent with the polls''. The polling results provides the null model from which data is simulated. Because each poll has a margin of error, this is used to simulate different scenarios that might have resulted on election day, if the polls were on target. A null data set is generated as a set of draws from normal distribution, with mean equal to the difference in poll percentage of the latest state poll results, and standard deviation equal to 2.5, approximating a margin of error of 5\%. These samples are plotted as electoral buildings, and the plot made with the results from the election is placed randomly among them in a lineup of size $5\times 4$. If the null hypothesis is true the actual data plot should look just like any of the other plots, and not be identified by an observer. Figure~\ref{fig:tower} shows a plot of the electoral building with added context information and labels.
 %A party needs 270 electoral votes or more to win the presidential election. 
%\begin{figure}[htbp] 
%   \centering
%   \includegraphics[width=0.8\linewidth]{electoral-5-13.pdf} 
%   \caption{Which plot looks the most different from the other plots?  Lineup \#1 - electoral building; another set of four lineups can be found in figs.~{fig:elect.2} through~{fig:elect.5}}
%   \label{fig:elect.1}
%\end{figure}

>>>>

The lineup protocol introduced in \citet{buja:2009} can be used to test the significance of \hh{graphical} findings in the exploratory data analysis process. This methodology is  part of what is called visual statistical inference, which has been developed further  and validated by \citet{majumder:2013} in simulation studies of head to head comparisons with conventional inference. One of the major contributions of  \citet{majumder:2013}  is to define the power of the visual test and provide methods to obtain the power for a particular lineup. It was observed that the power can be as good or better than that of a conventional test in some scenarios.

@@

Figure~\ref{fig:tower} displays a lineup of 20 plots where one of the plots is observed data, while the remaining 19 plots are showing  data generated under a null model. Which one of the 20 plots is the most different from the others? 
When asked this question, 12 \hh{out} of 72 observers picked the  plot \hh{of the observed data} (located in panel number $3^2+4$ in the lineup).  The corresponding $p$-value is \hh{0.0077}, indicating sufficient evidence to reject the null hypothesis. 
% , with reasons given being 'asymmetry' (36\%), 'trend' (26\%) or `outliers' (16\%).

\hh{How do we interpret this finding, though?} For that, we need to know the context of the data and we need to have more information about \hh{how the null plots were generated}. 
This particular example investigates the results from the 2012 US presidential election in comparison to  poll results just prior to the election. 
Although this example is more simplistic than most of the tests conducted to date, it will serve the purpose of illustrating the lineup protocol. The data consists of the difference in poll results between the two (major) presidential candidates, Obama and Romney, for all states. Each panel in Figure~\ref{fig:elect-1} shows an `electoral building' \citep{mosley2010} where each state in the union is represented by a rectangle. This difference is plotted horizontally, and the height of each box corresponds to the state's electoral votes. Color indicates party affiliation.  The null hypothesis is ``that the election results were consistent with the polls''. The polling results provides the null model from which data is simulated. \hh{A normal model with mean and standard deviation based on a poll and its margin of error is used to} simulate different scenarios that might have resulted on election day, if the polls were on target. A null data set is generated as a set of draws from this model. % normal distribution, with mean equal to the difference in poll percentage of the latest state poll results, and standard deviation equal to 2.5, approximating a margin of error of 5\%. 
These samples are plotted as electoral buildings, and the plot made with the results from the election is placed randomly among them in a lineup of size $5\times 4$. If the null hypothesis is true, the actual data plot should look  like any of the other plots, and not be identifiable by an observer. Figure~\ref{fig:tower} shows the plot of the electoral building with added context information and labels.
 %A party needs 270 electoral votes or more to win the presidential election. 
%\begin{figure}[htbp] 
%   \centering
%   \includegraphics[width=0.8\linewidth]{electoral-5-13.pdf} 
%   \caption{Which plot looks the most different from the other plots?  Lineup \#1 - electoral building; another set of four lineups can be found in figs.~{fig:elect.2} through~{fig:elect.5}}
%   \label{fig:elect.1}
%\end{figure}
